First Order Alternation

### 2018-06-08 quantifiers/draft.tex
ok

For two integers $0 \leq i \leq j$, we denote by $[i,j]$ the set
$\set{i,i+1,\ldots,j}$ and by $[i]$ the set $[1,i]$, where $[0]$ is
the empty set. To simplify the presentation, we consider two sorts
$\Data$ and $\Bool$, where $\Data$ ranges over an infinite domain and
$\Bool$ has the usual interpretation. The $\Data$ sort is equipped
with a finite set of function symbols $f : \Data^{\#(f)} \rightarrow
\Data$, where $\#(f)\geq0$ denotes the number of arguments (arity) of
$f$. When $\#(f)=0$, we say that $f$ is a constant, and denote by
$\top$ and $\bot$ the only constants of sort $\Bool$, representing the
true and false values, respectively.  For conciseness, in the
following we blur the distinction between the sort and function
symbols and their interpretations. A \emph{predicate} is a function
symbol $p : \Data^{\#(p)} \rightarrow \Bool$, denoting a relation of
arity $\#(p)$.

Let $\vars = \set{x,y,z,\ldots}$ be an infinite countable set of
variables, ranging over the interpretation of the $\Data$ sort. Terms
are either constants of sort $\Data$, variables or function
applications. The set of \emph{positive} formulae is defined by the
syntax below: 
\[\phi := t \teq s \mid t \not\teq s \mid p(t_1,\ldots,t_{\#(p)}) 
\mid \phi_1 \wedge \phi_2 \mid \phi_1 \vee \phi_2 \mid \exists x ~.~
\phi_1 \mid \forall x ~.~ \phi_1\] where $t,s,t_1,\ldots,t_{\#(p)}$
denote terms. We denote by $\fv{}{\phi}$ the set of free variables in
$\phi$. A \emph{sentence} is a formula $\phi$ in which each variable
occurs under the scope of a quantifier, i.e. $\fv{}{\phi} =
\emptyset$. The size $\len{\phi}$ is the number of symbols occuring in
$\phi$.

@@

@@

Intuitively, $a$ is the input event and $X$ are the input data values
that trigger the transition, whereas $q$ and $y_1,\ldots,y_{\#(q)}$
are the current control state and data values pertaining to that
state, respectively. Without loss of generality, we consider, for each
predicate $q \in Q$ and each input event $a \in \Sigma$, at most one
such rule (two or more rules can be joined using disjunction).

Given a predicate $q \in Q$ and a tuple of data values
$d_1,\ldots,d_{\#(q)}$, the tuple $q(d_1,\ldots,d_{\#(q)})$ is called
a \emph{configuration}\footnote{Note that a configuration is not a
  logical term since data values cannot be written in logic.}.
Because we consider an alternating model of computation, an automaton
can find itself, at the same time, in more than one
configuration. Typically, in the classical literature on alternating
machines, this set is finite. In our case, due to the presence of
first-order quantifiers ranging over an infinite data domain, we
consider that an automaton can be also in an infinite set of
configurations, all at once.

To formalize the execution semantics of automata, we relate sets of
configurations to models of first-order sentences, as follows. Given a
formula $\phi$ and a valuation of its free variables $\nu$, each
interpretation $\I \in \sem{\phi}_\nu$ corresponds to a set of
configurations $\cube{\I} \isdef \set{q(d_1,\ldots,d_{\#(q)}) \mid
  \tuple{d_1,\ldots,d_{\#(q)}} \in q^\I}$, called a \emph{cube}. For a
set $\mathcal{M}$ of interpretations, we define $\cube{\mathcal{M}}
\isdef \set{\cube{\I} \mid \I \in \mathcal{M}}$.

Given a word $w=(a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n) \in
\Sigma[X]^*$, an \emph{execution} of
$\A=\tuple{\Sigma,X,Q,\iota,F,\Delta}$ over $w$ is a (possibly
infinite) forest $\F_w = \set{T_1,T_2,\ldots}$, where each $T_i$ is
defined below:
\begin{compactitem}
\item $T_i$ is rooted in a configuration from some cube $C \in
  \cube{\sem{\iota}}$ --- since $\iota$ is satisfiable, such cube
  always exists,
%
\item all paths from the root to a leaf of $T_i$ have length $n$,
%
\item if $q(d_1,\ldots,d_{\#(q)})$ is a node on level $j \in [0,n-1]$
  in $T_i$, then its children form a cube $C \in
  \cube{\sem{\psi}_\eta}$, where \(q(y_1,\ldots,y_{\#(q)})
  \arrow{a_{j+1}(X)}{} \psi \in \Delta\) is a transition rule and
  $\eta$ is a valuation that assigns each variable $y_k$ the value
  $d_k$, for $k \in [1,\#(q)]$ and each variable $x \in X$ the value
  $\nu_{j+1}(x)$ --- because $\fv{}{\psi} \subseteq
  \set{y_1,\ldots,y_{\#(q)}} \cup X$, the values assigned by $\eta$ to
  variables outside $\set{y_1,\ldots,y_{\#(q)}} \cup X$ are not
  important for $\cube{\sem{\psi}_\eta}$.
\end{compactitem}
A forest $\F_w$ is said to be \emph{accepting} if and only if the
frontier of each tree $T \in \F_w$ is labeled with configurations
$q(d_1,\ldots,d_{\#(q)})$ such that $q \in F$ is a final predicate. In
this case, we say that $\A$ accepts $w$ and denote by $\lang{\A}$ the
set of words accepted by $\A$. In this paper, we answer the following
questions: \begin{compactenum}
\item \emph{closure}: given automata $\A_i =
  \tuple{\Sigma,X,Q_i,\iota_i,F_i,\Delta_i}$, for $i=1,2$, do there
  exist automata $\A_\cap$, $\A_\cup$ and $\overline{\A}_1$ such that
  $L(\A_\cap) = L(\A_1) \cap L(\A_2)$, $L(\A_\cup) = L(\A_1) \cup
  L(\A_2)$ and $L(\overline{\A}_1) = \Sigma[X]^* \setminus L(\A_1)$ ?
%
\item \emph{emptiness}: given an automaton $\A$, is $L(\A) =
  \emptyset$?
\end{compactenum}

@@

For a sequence of input events $a_1 \ldots a_n \in \Sigma^*$, we
consider the sequence of formulae \(\phi_0 \stackrel{a_1}{\Rightarrow}
\phi_1 \stackrel{a_2}{\Rightarrow} \ldots \stackrel{a_n}{\Rightarrow}
\phi_n\), where $\phi_0 = \iota$ and $\phi_{i+1}$ is obtained from
$\phi_i$ by replacing each predicate atom $q(t_1,\ldots,t_{\#(q)})$
from $\phi_i$ by the formula $\stamp{\psi}{i+1}[t_1, \ldots,
  t_{\#(q)}/y_1,\ldots,y_{\#(q)}]$, where $q(y_1,\ldots,y_{\#(q)})
\arrow{a_i(X)}{} \psi \in \Delta$ is the only transition rule that
applies, or by $\bot$, if no transition rule applies. The following
lemma formalizes the relation between symbolic execution and the
low-level configuration-based semantics of automata:

@@

\subsection{Configuration-based Semantics}

We give a low-level operational semantics of automata, in terms of
sets of configurations, with a conjunctive meaning. A (possibly
infinite) set $C$ of configurations is also called a
\emph{cube}. Intuitively, an automaton $\A$ is in a cube $C$ if it is
simultaneously in all configurations $q(d_1, \ldots, d_{\#(q)})$ from
$C$. A positive sentence $\phi$ containing at least one occurrence of
a predicate, defines a (possibly infinite) set of cubes $\sem{\phi}$,
defined next.

A \emph{context} $q[x_1,\ldots,x_k] = q(\delta_1, \ldots,
\delta_{\#(q)})$ is a configuration in which one or more variables may
occur, i.e. $\delta_i \in \Data^\I \cup \set{x_1,\ldots,x_k}$, for all
$i \in [\#(q)]$. For a set of data values $D \subseteq \Data^\I$, we
let $q[x_1,\ldots,x_{i-1},D,x_{i+1}, \ldots, x_k] =
\{q[x_1,\ldots,x_{i-1},d,x_{i+1},\ldots,x_k] \mid d \in D\}$ be a set
of contexts. Observe that $q[D]$ is a cube, when $k=1$.

>>>>

For two integers $0 \leq i \leq j$, we denote by $[i,j]$ the set
$\set{i,i+1,\ldots,j}$ and by $[i]$ the set $[0,i]$. To simplify the
presentation, we consider two sorts $\Data$ and $\Bool$, where $\Data$
ranges over an infinite domain and $\Bool$ has the usual
interpretation. The $\Data$ sort is equipped with a finite set of
function symbols $f : \Data^{\#(f)} \rightarrow \Data$, where
$\#(f)\geq0$ denotes the number of arguments (arity) of $f$. When
$\#(f)=0$, we say that $f$ is a constant, and denote by $\top$ and
$\bot$ the only constants of sort $\Bool$, representing the true and
false values, respectively. A \emph{predicate} is a function symbol
$p : \Data^{\#(p)} \rightarrow \Bool$, denoting a relation of arity
$\#(p)$ and we write $\preds$ for the set of predicates.

Let $\vars = \set{x,y,z,\ldots}$ be an infinite countable set of
variables, ranging over the interpretation of the $\Data$ sort. Terms
are either constants of sort $\Data$, variables or function
applications. The set of \emph{positive} first-order formulae is
defined by the syntax below:
\[\phi := t \teq s \mid t \not\teq s \mid p(t_1,\ldots,t_{\#(p)}) 
\mid \phi_1 \wedge \phi_2 \mid \phi_1 \vee \phi_2 \mid \exists x ~.~
\phi_1 \mid \forall x ~.~ \phi_1\] where $t,s,t_1,\ldots,t_{\#(p)}$
denote terms. We denote by $\fv{}{\phi}$ the set of free variables in
$\phi$. A \emph{sentence} is a formula $\phi$ in which each variable
occurs under the scope of a quantifier, i.e. $\fv{}{\phi} =
\emptyset$. The size $\len{\phi}$ is the number of symbols occuring in
$\phi$.

@@

Interpretations are partially ordered by the pointwise subset order,
defined as $\I_1 \subseteq \I_2$ if and only if $p^{\I_1} \subseteq
p^{\I_2}$ for each predicate $p \in \preds$. Given a set $\mathcal{M}$
of interpretations, a minimal element $\I \in \mathcal{M}$ is an
interpretation such that for no other interpretation $\I' \in
\mathcal{M} \setminus \set{\I}$ do we have $\I' \subseteq \I$. For a
formula $\phi$ and a valuation $\nu$, we denote by $\minsem{\phi}_\nu$
and $\minsem{\phi}$ the set of minimal interpretations from
$\sem{\phi}_\nu$ and $\sem{\phi}$, respectively.

@@

Intuitively, $a$ is the input event and $X$ are the input data values
that trigger the transition, whereas $q$ and $y_1,\ldots,y_{\#(q)}$
are the current control state and data values pertaining to that
state, respectively. Without loss of generality, we consider, for each
predicate $q \in Q$ and each input event $a \in \Sigma$, at most one
such rule --- two or more rules can be joined using disjunction.

The execution semantics of of automata is given in close analogy with
the case of boolean alternating automata, with transition rules of the
form $q \arrow{a}{} \phi$, where $q$ is a boolean constant and $\phi$
a positive boolean combination of such constants. For instance, $q_0
\arrow{a}{} q_1 \wedge q_2 \vee q_3$ means that the automaton can
choose to transition in either both $q_1$ and $q_2$ or in $q_3$
alone. This intuition leads to saying that the steps of the automaton
are defined by the minimal boolean models of the transition
formulae. In this case, both $\set{q_1 \leftarrow \top, q_2 \leftarrow
  \top, q_3 \leftarrow \bot}$ and $\set{q_1 \leftarrow \bot, q_2
  \leftarrow \bot, q_3 \leftarrow \top}$ are minimal models, but not
$\set{q_1 \leftarrow \top, q_2 \leftarrow \top, q_3 \leftarrow \top}$,
which is a model but is not minimal.

Given a predicate $q \in Q$ and a tuple of data values
$d_1,\ldots,d_{\#(q)}$, the tuple $q(d_1,\ldots,d_{\#(q)})$ is called
a \emph{configuration}\footnote{Note that a configuration is not a
  logical term since data values cannot be written in logic.}. To
formalize the execution semantics of automata, we relate sets of
configurations to models of first-order sentences, as follows. Given a
formula $\phi$ and a valuation of its free variables $\nu$, each
interpretation $\I \in \sem{\phi}_\nu$ corresponds to a set of
configurations $\cube{\I} \isdef \set{q(d_1,\ldots,d_{\#(q)}) \mid
  \tuple{d_1,\ldots,d_{\#(q)}} \in q^\I}$, called a \emph{cube}. For a
set $\mathcal{M}$ of interpretations, we define $\cube{\mathcal{M}}
\isdef \set{\cube{\I} \mid \I \in \mathcal{M}}$. Given a word
$w=(a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n) \in \Sigma[X]^*$, an
\emph{execution} of $\A=\tuple{\Sigma,X,Q,\iota,F,\Delta}$ over $w$ is
a (possibly infinite) forest $\F_w = \set{T_1,T_2,\ldots}$, where each
$T_i$ is defined below:
\begin{compactitem}
\item $T_i$ is rooted in a configuration from some cube from
  $\cube{\minsem{\iota}}$ --- because $\iota$ is satisfiable, such cube
  always exists,
%
\item all paths from the root to a leaf of $T_i$ have length $n$, and
%
\item if $q(d_1,\ldots,d_{\#(q)})$ is a node on level $j \in [0,n-1]$
  in $T_i$, then its children form a cube from
  $\cube{\minsem{\psi}_\eta}$, where \(q(y_1,\ldots,y_{\#(q)})
  \arrow{a_{j+1}(X)}{} \psi \in \Delta\) is a transition rule and
  $\eta$ is a valuation that assigns each variable $y_k$ the value
  $d_k$, for $k \in [1,\#(q)]$ and each variable $x \in X$ the value
  $\nu_{j+1}(x)$ --- because $\fv{}{\psi} \subseteq X \cup
  \set{y_1,\ldots,y_{\#(q)}}$, the values assigned by $\eta$ to
  variables outside $X \cup \set{y_1,\ldots,y_{\#(q)}}$ have no impact
  on $\cube{\minsem{\psi}_\eta}$.
\end{compactitem}
A forest $\F_w$ is said to be \emph{accepting} if and only if the
frontier of each tree $T \in \F_w$ is labeled with configurations
$q(d_1,\ldots,d_{\#(q)})$ such that $q \in F$ is a final predicate. In
this case, we say that $\A$ accepts $w$ and denote by $\lang{\A}$ the
set of words accepted by $\A$. In this paper, we answer the following
questions: \begin{compactenum}
\item \emph{closure}: given automata $\A_i =
  \tuple{\Sigma,X,Q_i,\iota_i,F_i,\Delta_i}$, for $i=1,2$, do there
  exist automata $\A_\cap$, $\A_\cup$ and $\overline{\A}_1$ such that
  $L(\A_\cap) = L(\A_1) \cap L(\A_2)$, $L(\A_\cup) = L(\A_1) \cup
  L(\A_2)$ and $L(\overline{\A}_1) = \Sigma[X]^* \setminus L(\A_1)$ ?
%
\item \emph{emptiness}: given an automaton $\A$, is $L(\A) =
  \emptyset$?
\end{compactenum}

@@

For a sequence of input events $a_1 \ldots a_n \in \Sigma^*$, we
consider the sequence of formulae \(\phi_0 \stackrel{a_1}{\Rightarrow}
\phi_1 \stackrel{a_2}{\Rightarrow} \ldots \stackrel{a_n}{\Rightarrow}
\phi_n\), where $\phi_0 = \iota$ and, for each $i \in [n-1]$, the
formula $\phi_{i+1}$ is obtained from $\phi_i$ by replacing each
predicate atom $q(t_1,\ldots,t_{\#(q)})$ from $\phi_i$ by the formula
$\stamp{\psi}{i+1}[t_1, \ldots, t_{\#(q)}/y_1,\ldots,y_{\#(q)}]$,
where $q(y_1,\ldots,y_{\#(q)}) \arrow{a_i(X)}{} \psi \in \Delta$ is
the only transition rule that applies, or by $\bot$, if no transition
rule applies. The following lemma formalizes the relation between the
symbolic and the configuration-based semantics of automata:

@@
